---
title: "Life Expectancy Dynamics"
subtitle: "A Cross Cultural Perspective"
author: "Esther Crespo"
date: "04/25/2024"
format:
  html: 
   theme: flatly
   backgroundcolor: lightgrey
   toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

#libraries
library(tidyverse)
library(maps)
library(ggplot2)
library(plotly)
library(gapminder)
library(dplyr)
library(RColorBrewer)

#data
Unicef_ind_1 <- read.csv("D:/Libraries/DCU UNIVERSITY COURSES/DATA ANALITYCS/unicef_indicator_1 (1).csv")
Unicef_ind_metadata <- read.csv("D:/Libraries/DCU UNIVERSITY COURSES/DATA ANALITYCS/unicef_metadata (1).csv")

#transformations

Join_data <- left_join(Unicef_ind_metadata, Unicef_ind_1, by = c("year" = "time_period", "country","alpha_2_code", "alpha_3_code"))
gapminder <- select(gapminder, country, continent)


Join_data <- left_join(Join_data, distinct(gapminder), by = "country")
```

This study delves into *Africa, Asia, and the Americas*, investigating nations where the female adolescent population, aged 15-19, exhibits notably higher life expectancy within marriage or union. Through an analysis, factors such as GDP per capita, cultural norms, religious influences, and educational opportunities will be studied to better comprehend the implications of this trend.By examining these variables, the aim to shed light on the socio-economic, cultural, and educational dynamicsthat contribute to elevated life expectancy among adolescent girls in certain regions.

This map shows a visual representation of life expectancy across the globe in 2005, with countries/regions with higher life expectancy appearing in different colors compared to those with lower life expectancy.

## The world since the year 2000 and 2005.

```{r}

#map 1

map_world <- map_data("world")
data_join_2005 <- Join_data %>%
  filter(year == 2005)

map_data_join_2005 <- full_join(data_join_2005, map_world, by = c("country" = "region"))

 ggplot(data = map_data_join_2005) +
  aes(x = long, y = lat, group = group, fill = Life.expectancy.at.birth..total..years.) +
  geom_polygon() +
  scale_fill_gradient(low = "pink", high ="darkblue", na.value ="grey")+
  labs(
    title = "LIFE EXPECTANCY IN 2005 ACROSS THE WORLD",
    subtitle = "Countries in grey have no data",
    caption = "Source: R package - Unicerf_data",
    x = "Longitude",
    y = "Latitude",
    fill = "Life expectancy",
    
) + 
  theme_bw()
 
 #| fig-cap: "This map shows a visual representation of life expectancy across the globe in 2005, with countries/regions with higher life expectancy appearing in different colors compared to those with lower life expectancy. "
```

**Child marriage in countries like India remains a significant societal challenge with far-reaching implications, particularly concerning population dynamics**. This practice not only deprives children, especially girls, of their basic rights and opportunities but also contributes to perpetuating cycles of poverty and inequality.

```{r}
#map 3

Join_data %>%
  filter(year == 2000)  %>%
  full_join(map_world, by = c("country" = "region"))  %>%
  ggplot(aes(x = long, y = lat, group = group, fill = Population..total)) +
  geom_polygon()+
  scale_fill_gradient(low = "green", high ="red", na.value ="grey")+
  labs(
    title = "POPULATION DISTRIBUTION IN THE YEAR 2000",
    subtitle = "Countries in grey have no data",
    caption = "Source: R package - Unicerf_data",
    x = "Longitude",
    y = "Latitude",
    fill = "Total Population",
    
) + 
  theme_bw()
  options(scipen = 999)
  
  #| fig-cap: " This map snippet retrieves population data for the year 2000, joins it with geographic data, and each country/region is represented by a polygon colored  based on its population. Countries with higher populations will appear in darker colors compared to those with lower populations."


```

During this period, advancements in healthcare, sanitation, nutrition, and public health policies have led to significant improvements in life expectancy globally. Many countries have witnessed a steady increase in life expectancy, reflecting improved living conditions, better access to healthcare services, and advancements in medical technology. This upward trend signifies the success of public health interventions and the resilience of human populations in combating diseases and improving overall well-being.

## Evolution over the years - Time series visualization

```{r}
#time series visualization 1


timeseries_plot_1 <- Join_data %>%
  ggplot () +
  aes(year, Life.expectancy.at.birth..total..years., group = country, color = continent) +
  geom_line()+
  labs(
    title = "Life expectancy Distribution over the years (1960-2022)",
    subtitle = "Countries in grey have no data",
    caption = "Source: R package - Unicerf_data",
    x = "Year",
    y = "Life expectancy",
    fill = "Continent",
    
) + 
  theme_bw()

ggplotly(timeseries_plot_1)


#| fig-cap: "This time series visualization snippet likely creates a line chart where the x-axis represents years, the y-axis represents life expectancy, and each line represents a specific country. The lines are colored based on the continent to which each country belongs. This visualization allows to see trends in life expectancy over time for different countries, grouped by continent. "
```

**AFRICA**: This continent showcases higher prevalence of countries such as Niger, Mali, South Sudan, and Mozambique with elevated percentages of child marriage or in union. From 40% to 64% of its population experiment this issue.

**ASIA**: Countries such as Bangladesh, with a percentage as high as 32.86%, have a significant portion of children experiencing early marriage. This is followed by countries like Iran, Iraq, Pakistan, Bhutan, Nepal, Afghanistan, Laos, and Papua-Guinea

**AMERICA**: Suriname stands out as the country with a highest percentage of girls married (34.40%), while neighbouring nations such as Venezuela, Colombia, Paraguay, Ecuador, Honduras, Nicaragua, Guatemala, Panama, Belize and Dominican Republic exhibit rates ranging between 13% and 23% of girls married.

```{r}
#timeseries 2



timeseries_plot_2 <- Join_data %>%
  ggplot () +
  aes(year, Population..total, group = country, color = continent) +
  geom_line()+
  labs(
    title = "Population Distribution over the years (1960-2022)",
    subtitle = "Countries in grey have no data",
    caption = "Source: R package - Unicerf_data",
    x = "Year",
    y = "Total Population",
    fill = "Total Population",
    
) + 
  theme_bw()

ggplotly(timeseries_plot_2)
 #| fig-cap: "The main difference between this code and the previous one is the variable used on the y-axis. Here, it's the total population. This creates a visualization focused on population trends. "
```

Observationally, countries with lower levels of GDP per capita tend to grapple more with the issue of child marriage. This phenomenon is largely attributed to poverty-related factors and limited access to education within the nations. As observed previously, the continents most affected by that issue are Africa, Asia, and South America.

*Lower GDP* typically correlates with levels of poverty. Families living in poverty may see child marriage as a coping mechanism to alleviate financial strain by reding the number of mouths to feed or receiving dowry payments.

Countries such as *Suriname*, *Bangladesh*, *Niger*, *Mali*, *South Sudan*, and *Mozambique* may struggle to provide economic opportunities for their citizens, particularly in rural areas. In such as contexts, families may view marriage to secure economic stability or forge alliances with wealthier families. Also, these countries tend to have higher levels of inequality and women have limited autonomy. This perpetuates the practice of child marriage as girls are not given the same opportunities for education, employment, or decision-making as boys.

Traditional and cultural norms regarding gender roles and marriage are often deeply entrenched in lower GDP countries.

## Evolution over the years - Scatter plot visualization

```{r}

#scatter plot


Join_data %>%
  filter(year >= 2005) %>%
  ggplot() +
  aes(GDP.per.capita..constant.2015.US.., Life.expectancy.at.birth..total..years., color = continent, size = Population..total ) +
  geom_point(alpha = 0.2) +
  facet_wrap(~ year, nrow = 1) +
  scale_x_continuous(
    limits = c(0, 5000000),
    breaks = c(2000000, 4000000),
    labels = scales::unit_format(unit = "K", scale = 0.001) 
  ) +
  labs(
    x = "obs_value",
    y = "Life Expectancy",
    title = "Evolution of life expectancy from 2005 to 2023 per continent."
  ) +
  guides(color = "none", size = "none") +
  theme_classic() +
  theme(text = element_text(family = "serif"))
  #| fig-cap: " This visualization snippet creates a scatter plot with several points for each year (from 2005 onwards). Each point represents a country, positioned      based on its GDP per capita (x-axis) and life expectancy (y-axis). The color of the point indicates the continent, and the size of the point represents the population   of the country. It is possible to see how these relationships might change over time across different years."

```

According to the provided data, 2012 recorded the highest average levels of GDP per capita globally. Predominantly, countries in Asia exhibit the highest values, closely followed by those in the Americas. Notably, in 2020, Europe emerges as a standout region in terms of GDP per capita.

## Evolution over the years - Bar chart visualization

```{r}
#bar chart

#| fig-cap: "This visualization shows the average life expectancy trends across continents over time. By using a bar chart with reordered continents, we can easily compare how average life expectancy changes for different continents across the years (from 2005 onwards."

Join_data %>%
  filter(year >= 2005) %>%
  group_by(continent, year) %>%
  summarise(m_Life.expectancy = mean(Life.expectancy.at.birth..total..years., na.rm = TRUE)) %>%
  drop_na() %>%
  ggplot()+
  aes(reorder(continent,m_Life.expectancy),m_Life.expectancy, fill = continent ) + 
  geom_col() + 
  facet_wrap(~ year, nrow = 1)+
  labs(
    x = "",
    y = "Life expectancy",
    fill = "Continent",
    title = "Evolution of life expectancy from 2005 to 2023 per country."
) + 
  theme_classic() +
  theme(
    text = element_text(family = "serif"),
    axis.text.x = element_blank()
  ) +
  scale_fill_manual(values = c("red", "yellow", "darkblue", "green", "pink"))
```

## Implications and recommendations

The implications of child marriage are profound and far reaching, affecting not only girls but also their families, communities, and societies.

-Child brides face significant health risks, including complications and childbirth due to their young age. They are more likely to experience maternal mortality and other pregnancy-related complications. They might suffer from poor-nutrition and mental health issues from the trauma of early marriage.

-Poorest countries like *Bangladesh* or *Suriname* in South America, present high levels of inequality among girls and boys. The negative outcome of child marriage extends beyond the individual girls to their children. Children born to young mothers are more likely to experience poor health outcomes, malnutrition and reduced access to educations and healthcare services.

-Especially in the countries previously addressed, efforts should be focus on empowering girls with education, economic opportunities and implementing legal and policy frameworks to protect their rights and prevent early marriage.
